Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-08-07 manuscripts/reply/reviews/response.tex

%&<latex>
\documentclass[12pt]{article}

\usepackage{parskip}
\input{../../utils/preamble.tex}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{enumitem}
\input{../../utils/macros.tex}

\newcounter{commentCounter}
\newcommand{\revcomment}[1]{{\addtocounter{commentCounter}{1}}
    \medskip \hrule \noindent
\textbf{\arabic{section}.\arabic{commentCounter}}: {\sl #1}\par\xspace}
\newcommand{\response}[1]{{\addtolength{\leftskip}{0.25in} #1\par}\xspace}

\let\quoteOld\quote
\let\endquoteOld\endquote
\renewenvironment{quote}{\sffamily\small\quoteOld}{\endquoteOld}

\begin{document}
\setlength{\parindent}{0in}

My co-authors and I greatly appreciate the helpful comments of the editors and
reviewers.  In this document, I provide our responses to these comments,
including descriptions of the changes we have made.

\section{Editor's Comments}\label{ed}
\setcounter{commentCounter}{0}
\revcomment{
    In particular, the paper needs to be more focused and would benefit from
    avoiding unnecessary polemical considerations. Importantly, the core
    arguments could be made in much fewer words, given previous publications. A
    resubmission should be cut by at least 25\% in the main text. Similarly
    consider removing 1 or 2 figures (perhaps fig. 5 and 6).
}
\response{
    We have reduced the number of words in the main text by a little more than
    25\%, and moved Figures 5 and 6 to the Supporting Information.
    We have re-written many sections to make the paper less polemical and more
    focused.
}

@@

\section{Associate Editor's Comments}\label{ed}
\setcounter{commentCounter}{0}
\revcomment{
    The most negative review suggests that the authors (and Hickerson et al)
    might have reached a point of diminishing returns, where further exchange
    in the pages of Evolution might not be informative for our readership. My
    own view is somewhat in line with this reviewer. I do think there are some
    new ideas in the present manuscript, and I also want to honor the previous
    AE who solicited an additional response from Oaks and colleagues to the
    last Hickerson et al. paper. But I would also say that after this paper,
    enough is enough, in my opinion \ldots
}
\response{
    We agree. We are very appreciative of this opportunity to clarify some
    aspects of Hickerson et al.'s response, and feel our reply will make a nice
    end to an interesting discussion about Bayesian model selection.

    We have re-written a lot of the manuscript and feel the new version will be
    of greater interest to readers interested in Bayesian selection of
    evolutionary models.
}

@@

\revcomment{
     I also echo the suggestion from reviewer 2 to shorten this paper as much
     as is practical, especially by moving things to the supplement. One can
     safely assume that readers still following the thread have read the
     preceding papers in the series, I think.
}
\response{
    We are greatful for this insight by the editor and reviewers that our
    paper was too long and lacked focus.
    We greatly reduced the length of the main text, moved some items (including
    2 figures) to the supplement, and re-wrote a lot of the manuscript to focus
    on the most interesting aspects of Bayesian model choice brought to
    light by this series of papers.
}

\revcomment{
    One small point: there is a typo on line 19: Oaks (2014) should be (Oaks
    2014).
}
\response{
    Corrected. Thank you.
}

\section{Reviewer 1's Comments}\label{rev1}
\setcounter{commentCounter}{0}
\revcomment{
    At this point, I think at least for the readers of Evolution, we are
    reaching the point of diminishing returns and perhaps the discussion has
    burned out. The arguments of how broad priors or mixes of specific and
    diffuse priors yield poor posteriors is indeed important but perhaps
    another venue would be appropriate for this discussion---also with a
    greater emphasis on solutions, if they exist for non-model organisms, to
    common problems affecting most molecular ecologists (particularly regarding
    substitution rate uncertainty).

}
\response{
    We have re-written much of the manuscript in order to make it of greater
    interest to the broad readership of \emph{Evolution}.
    We have added emphasis of solutions to the problem in several places in the
    manuscript (abstract, introduction, and conclusions). The solution was the
    focus of a different paper \citep{Oaks2014dpp} that introduces an improved
    method for estimating shared divergence histories. We now draw attention to
    this positive outcome.
}

@@

\revcomment{
    The paper is rather long for simply arguing against a model averaging
    approach. I recommend to move the coin example to the supp info, as this is
    a simple text book example that does require to be published in a journal
    such as evolution, but is still a nice illustration so that it should be
    published as supp info to the paper. Also, I think the paper sometimes
    reads like a direct accusation of the findings in Hickerson et al, to agree
    to which they might be more emotional than required for the presented
    arguments. Paragraphs such as L96-100, L104-108, L222-237 or L287-289
    (among others later in the text and, especially, in the supp info) for
    instance, could simply be omitted or shortened, which would make reading
    the paper much more enjoyable, as it would focus on the valid arguments,
    instead of highlighting subtleties only personally important to people
    included in the debate.

}
\response{
    We thank Reviewer 2 for this insight. We agree that the paper was too long
    and lacked focus. Reviewer 2 also points out several sections that might be
    interpreted as emotionally charged or polemical, which we appreciate. We
    did not intend for our response to read as an ``attack'' of the work of
    Hickerson et al., but Reviewer 2 helped us to realize that wording in a
    response paper needs to be carefully chosen to avoid this interpretation.

    We have moved the coin-flipping example to the supplement.

    In regard to the sections specifically highlighted by Reviewer 2:
    \begin{description}[align=left,labelindent=0.5in,leftmargin=0.75in]
        \item[L96-100 and L104-108]
            We have removed this section.
        \item[L222-237]
            We see how this paragraph can be interpreted as polemical. However,
            this brief summary of the methods and results of Hickerson et al.'s
            model-averaging approach is needed to set up the motivation for our
            subsequent analyses.
        \item[L287-289]
            We agree this section was superfluous and removed it.
    \end{description}

@@

    We have also removed and edited many other sections of the paper to try to
    eliminate sections that might seem overly negative. Specifically, we
    removed all quotations of sections from Hickerson et al. While we only
    intended to accurately represent their positions, we felt this is too
    easily interpreted as contentious.
}

\revcomment{
    While criticizing the approach proposed by Hickerson et al. and admitting
    the power problem of msBayes, the paper falls short of providing much of an
    alternative. However, I do think that there could be other ways of solving
    the problem of too many parameters for few summary statistics (section 6).
    This is especially true since ABC is known to be particularly sensitive to
    the choice of summary statistics for model choice (see Roberts et al. PNAS
    2011). I think the paper would be much stronger, if it could also discuss
    (but not necessarily test or implement) potential solutions to the apparent
    problem.
}
\response{
    This is a great point, and we have added emphasis on solutions to the
    problem, as recommended by Reviewer 2. As mentioned above, solutions to
    this problem were the focus of a separate paper, and were too extensive to
    discuss in detail in the response paper. However, we now highlight the main
    points of this paper to make the response more positive, and to to refer
    readers to an improved methodology.
}

@@

\revcomment{
    One alternative suggestion could be to formulate the problem as a
    hierarchical model, in which one seeks only to infer mean and variance in
    divergence times. Such a model would be much easier to fit, as the summary
    statistics ordering problem discussed (and solved) in the original msBayes
    paper could be circumvented easily, as the model would not seek to infer
    taxon specific divergence times. Also, it could then be compared to a model
    of zero variance using either Bayes factors, or by assessing the posterior
    probability of the variance to be zero using a simulation / bootstrap
    approach. If a model of co-divergence is preferred, the inferred posterior
    on the mean divergence would directly serve as an estimate of the joint
    divergence time. If independent divergences are inferred, then a simpler
    model of a single taxon pair could be used to infer the divergence times of
    each taxon pair individually, as they were found to be independent. For
    such a model, the same ABC simulations could be reused for all pairs,
    making this step computationally very cheap. But I'm sure there would also
    be other alternatives.
}
\response{
    These are great points raised by Reviewer 2. Many of these are
    discussed in detail in a separate paper \citep{Oaks2014dpp}.
    We highlight the improved methodology presented there to add a positive
    note to the response.
}

\revcomment{
    A minor issue on L53: while it is true that all numerical methods suffer
    from large prior ranges, it is also true that effect is pronounced in ABC,
    as the sampling occurs from the prior, and not the posteriors, as is done
    in techniques such as MCMC. Hence, it is not a problem that is completely
    extrinsic to the ABC machinery.
}
\response{
    We thank Reviewer 2 for pointing out this section, which was admitedly very
    confusing. We have re-written this section (and a lot of the manuscript) to
    try to make this distinction clearer. Our goal is to distinguish two
    different ways that broad uniform priors could cause the poor behavior of
    msBayes that was shown in the original Oaks et al. paper.  We now pose
    these as alternative hypotheses to explain the method's tendency to support
    clustered divergence times. We also discuss the distinct predictions made
    by each hypothesis, and how best to improve the method under each of them.

### 2014-08-07 manuscripts/reply/cover-letter/cover-letter.tex
Updating reply cover letter.
More work on reply ms and response to reviews.

\documentclass[letterpaper]{letter}
\usepackage{anysize}
\marginsize{1in}{1in}{0.5in}{0.5in}
\usepackage{setspace}
\usepackage{url}
\usepackage{verbatim}
\usepackage{parskip}
\usepackage{hyperref}
\hypersetup{pdfborder={0 0 0}, colorlinks=true, urlcolor=blue}

@@

\signature{Jamie Oaks \\
                 Postdoctoral Fellow \\ 
                 Department of Biology \\
                 University of Washington \\ 
                 Box 351800 \\ 
                 Seattle, WA 98195 \\ 
                 USA \\
                 \href{mailto:joaks1@uw.edu}{\texttt{joaks1@uw.edu}} \\ 
                 \href{http://www.phyletica.com}{\texttt{www.phyletica.com}}}
%\address{Biodiversity Institute \\ University of Kansas \\ Dyche Hall, 1345 Jayhawk Blvd \\ Lawrence, KS 66045}
\begin{document}
\begin{letter}{The Society for the Study of Evolution \\
                     PO Box 299 \\
                     St. Louis, MO 63166-0299}
\opening{Dear Editor:}
% \raggedright
I am submitting a manuscript entitled ``Implications of uniformly distributed,
empirically informed priors for phylogeographical model selection: A reply to
Hickerson et al.'' to be considered for publication in the \emph{International
Journal of Organic Evolution}.
This manuscript is a response to a paper by Hickerson et al.\ (2014,
\emph{Evolution} 68:284--294), which was a response to a paper published last
year by my co-authors and me (Oaks et al.\ 2013, \emph{Evolution}
67:991--1010).
Former Associate Editor Dr.\ Laura Kubatko invited us to write this response,
with the intention of handling its review.
The following is an excerpt from Dr.\ Kubatko's comments in the decision letter
for Hickerson et al.\ (2014):

\begin{quote}
Thus, I propose the following: Hickerson et al. should prepare and
submit the final version of their response.  I urge them to fully
consider all of the points raised by Oaks in his most recent review of their
paper. However, since the point of view of Hickerson et al. seems to
be at odds with Oaks et al., I would like to give Oaks et al. the opportunity
to respond, if they so choose. I have therefore recommended acceptance
for this version of the manuscript, and I ask that Hickerson et al. provide a
copy of the final version to Jamie Oaks once they have submitted it, so
that he and his co-authors can consider whether they wish to respond.
\end{quote}

In our response, we use simulated and biological data to evaluate the empirical
Bayesian method proposed by Hickerson et al.\ (2014) as a solution to the
problems revealed by Oaks et al.\ (2013) regarding the power and accuracy of the
comparative phylogeographical method implemented in the software
\texttt{msBayes}.
We demonstrate their approach does not mitigate the method's bias toward
supporting models with less parameter space, but rather allows it to manifest
in strong preference for models that often exclude from consideration the true
values of the model's parameters.
Our results also strongly suggest the underlying statistical cause of the
method's tendency to support simple models.
More generally, we discuss the theoretical and practical issues of empirical
Bayesian approaches to model choice; these issues are important for empiricists
applying methods of Bayesian model selection to carefully consider.
Lastly, we discuss ways to mitigate the biases of the \texttt{msBayes} method
while avoiding negative side effects of empirically informed uniform priors.

We performed the work described in the manuscript following the principles of
Open Notebook Science; using version-control software, we make progress in all
aspects of the work freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce my results is provided there.

Thank you for your consideration, and I look forward to your correspondence
regarding this manuscript.

>>>>

\documentclass[letterpaper]{letter}
\usepackage{anysize}
\marginsize{1in}{1in}{0in}{0.5in}
\usepackage{setspace}
\usepackage{url}
\usepackage{verbatim}
\usepackage{parskip}
\usepackage{hyperref}
\hypersetup{pdfborder={0 0 0}, colorlinks=true, urlcolor=blue}

@@

\signature{Jamie Oaks \\
                 Postdoctoral Fellow \\ 
                 Department of Biology \\
                 University of Washington \\ 
                 Box 351800 \\ 
                 Seattle, WA 98195 \\ 
                 USA \\
                 \href{mailto:joaks1@uw.edu}{\texttt{joaks1@uw.edu}} \\ 
                 \href{http://www.phyletica.com}{\texttt{www.phyletica.com}}}
%\address{Biodiversity Institute \\ University of Kansas \\ Dyche Hall, 1345 Jayhawk Blvd \\ Lawrence, KS 66045}
\begin{document}
\begin{letter}{The Society for the Study of Evolution \\
                     PO Box 299 \\
                     St. Louis, MO 63166-0299}
\opening{Dear Editor:}
% \raggedright
I am re-submitting a manuscript (previous submission ID 14-0137) entitled
``Implications of uniformly distributed, empirically informed priors for
phylogeographical model selection:
A reply to Hickerson et al.'' to be considered for publication in the
\emph{International Journal of Organic Evolution}.
This manuscript is a response to a paper by Hickerson et al.\ (2014,
\emph{Evolution} 68:284--294), which was a response to a paper published last
year by my co-authors and me (Oaks et al.\ 2013, \emph{Evolution}
67:991--1010).
Former Associate Editor Dr.\ Laura Kubatko invited us to write this response.

In the manuscript, we highlight alternative hypotheses for why the popular
comparative phylogeographcial method, \texttt{msBayes}, often incorrectly
infers simultaneous divergence across co-distributed species.
Using analyses of simulated and empirical data, we test the distinct
predictions of these hypotheses and find the cause of the bias is not
approximation error, but rather strongly weighted marginal likelihoods of
divergence models under uniform priors on divergence times.
We also evaluate the empirical Bayesian method proposed by Hickerson et al.\
(2014) as a solution to the problems revealed by Oaks et al.\ (2013) regarding
the power and accuracy of the method.
We demonstrate their approach does not mitigate the method's bias toward
supporting models with less parameter space, but rather allows it to manifest
in strong preference for models that often exclude from consideration the true
values of the model's parameters.
More generally, we discuss the theoretical and practical issues of empirical
Bayesian approaches to model choice; these issues are important for empiricists
applying methods of Bayesian model selection to carefully consider.
We highlight ways to mitigate the biases of the \texttt{msBayes} method while
avoiding negative side effects of empirically informed uniform priors.
Overall, this paper highlights the importance of choosing prior assumptions
about parameters in Bayesian selection of models of evolution; if our prior
assumptions place too much weight in unlikely regions of parameter space, the
true posterior will often support the wrong model of evolutionary history.

We performed the work described in the manuscript following the principles of
Open Notebook Science; using version-control software, we make progress in all
aspects of the work freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce our results is provided there.

The previous submission of this manuscript was handled by Associate Editor Dr.\
Luke Harmon who obtained three anonymous reviews.
We have incorporated the helpful revisions recommended by Dr.\ Harmon and the
reviewers.
My point-by-point response to their comments follows this letter.
Thank you for your consideration, and I look forward to your correspondence
regarding this manuscript.

### 2014-08-07 manuscripts/reply/reply-abstract.tex
More work on reply ms and response to reviews.

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that a common process affected
the populations.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such a pattern of simultaneous
divergence across taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa when prior uncertainty about divergence times
is represented by a uniform distribution.
However, the papers differ about the best explanation and solution for this
problem.
\citet{Hickerson2013} suggested the lack of robustness in \msb analyses is
caused by numerical approximation error due to the inefficiency of the method's
rejection algorithm under broad uniform priors on divergence times
(Hypothesis~\ref{hypError}).
To overcome this, they proposed a model-averaging approach that uses narrow,
empirically informed uniform priors.
\citet{Oaks2012} suggested the method's behavior was due to the sensitivity of
marginal likelihoods to the uniformly distributed priors on divergence times
(Hypothesis~\ref{hypML}), and proposed alternative prior probability
distributions to mitigate this sensitivity.
As predicted by fundamental principles of Bayesian model choice, more flexible
distributions that accommodate prior uncertainty about parameters without
placing excessive weight in vast regions of parameter space with low
likelihood increase the method's robustness and power to detect temporal
variation in divergences \citep{Oaks2014dpp}.
Here, we use analyses of simulated and empirical data to show that the tendency
of \msb analyses to support models of shared divergences is primarily due to
Hypothesis~\ref{hypML}, whereas Hypothesis~\ref{hypError} is an untenable
explanation for the bias.
We also demonstrate the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} will often erroneously support shared divergence events
and is dangerous in the sense that the empirically-derived uniform priors often
exclude from consideration the true values of the divergence-time parameters.
All of these papers have highlighted the importance of choosing prior
assumptions about parameters in Bayesian model selection.
If our prior assumptions place too much weight in unlikely regions of parameter
space such that the true posterior supports the wrong model of evolutionary
history, no amount of computation can rescue our inference.

>>>>

Establishing that a set of population splitting events occurred at the same
time can be a potentially persuasive argument that a common process affected
the populations.
Recently, \citet{Oaks2012} assessed the ability of an approximate-Bayesian
model choice method (\msb) to estimate such a pattern of simultaneous
divergence across taxa, to which \citet{Hickerson2013} responded.
Both papers agree that the primary inference enabled by the method is very
sensitive to prior assumptions and often erroneously supports shared
divergences across taxa when prior uncertainty about divergence times
is represented by a uniform distribution.
However, the papers differ about the best explanation and solution for this
problem.
\citet{Hickerson2013} suggested the lack of robustness in \msb analyses is
caused by numerical approximation error due to the inefficiency of the method's
rejection algorithm under broad uniform priors on divergence times
(Hypothesis~\ref{hypError}).
To overcome this, they proposed a model-averaging approach that uses narrow,
empirically informed uniform priors.
\citet{Oaks2012} suggested the method's behavior was due to the sensitivity of
marginal likelihoods to the uniformly distributed priors on divergence times
(Hypothesis~\ref{hypML}), and proposed alternative prior probability
distributions to mitigate this sensitivity.
Here, we use analyses of simulated and empirical data to show that the tendency
of \msb analyses to support models of shared divergences is primarily due to
Hypothesis~\ref{hypML}, whereas Hypothesis~\ref{hypError} is an untenable
explanation for the bias.
If our prior assumptions place too much weight in unlikely regions of parameter
space such that the true posterior supports the wrong model of evolutionary
history, no amount of computation can rescue our inference. 
We also demonstrate the empirical Bayesian model-averaging approach of
\citet{Hickerson2013} is dangerous in the sense that the empirically-derived
uniform priors often exclude from consideration the true values of the
divergence-time parameters.
Furthermore, this approach will often strongly support models of highly
clustered divergence times when analyzing data from taxa that diverged randomly
over millions of generations.
As predicted by fundamental principles of Bayesian model choice, more flexible
distributions that accommodate prior uncertainty about parameters without
placing excessive weight in vast regions of parameter space with low
likelihood increase the method's robustness and power to detect temporal
variation in divergences \citep{Oaks2014dpp}.
